\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-21 1-intro.Rmd
intro done

Spatio-temporal data concerns variables observed in the space across time and this paper focuses on the spatio-temporal vector data with fixed locations and regular time interval. This type of data can naturally been considered as a data cube and here a cube with three axes: spatial identifier, temporal identifier, and variable can be used to conceptualise spatio-temporal data. In this cubic formation, space only occupy one dimension, rather than two, to avoid the needs of hypercube for multivariate spatio-temporal data.

While conceptually the data format is relatively simple, the actual data object that comes to the analysts in the analysis can be in various forms. Figure \ref{fig:illu-input} shows four examples of wild spatio-temporal data: 1) combines the time invariant and varying variables into a single table through duplicating the time invariant variables; 2) splits those variables into a separate spatial table and a temporal table; 3) ...; and lastly 4) is array-based data, like NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

```{r illu-input, echo = FALSE, out.width = "100%", fig.cap="An illustration of different spatio-temporal data format: (1) a single table for all the variables; (2) separate tables for spatial and temporal variables; (3) a table of metadata used to query the database and a separate table for queried data; and (4) an array data structure.", fig.align='center', out.height = "40%"}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

@@

Different authors have proposed various data objects to work with spatial temporal data in \proglang{R} and this includes: \pkg{spacetime} [@spacetime] is built on the \pkg{sp} [@sp] and \pkg{xts} [@xts] class and both of which have been replaced by more recent spatial and temporal data objects. \pkg{spatstat} [@spatstat] implements a `ppp` class for analysing point pattern data; and more recent, \pkg{stars} [@stars] has implemented an spatio-temporal array for wrangling both vector and raster data.

\textcolor{red}{Think again on the motivation and why it is hard to work existing tools.} In spatio-temporal data analysis, analysts will receive one of the various data formats depends on the data provider and need to manually rearrange the data to fit into one of the data object in \proglang{R} before starting any analysis on the data itself.

Looking at spatial or temporal data analysis, data objects like \pkg{sf} [@sf] and \pkg{tsibble} [@tsibble] have smoothed the data analysis in these two domains while there has not yet been a spatio-temporal data structure that makes it easy to explore spatio-temporal data and this creates a gap in the software development. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: climate observations from weather stations, air quality variables from air monitoring stations, or any variables that are measured at fixed location across time.

@@

This paper describes the implementation of a new spatio-temporal data structure: \pkg{cubble}. \pkg{Cubble} implements a data structure that uses two forms to manage the switch between spatial and temporal dimension. With this structure, users can manipulate the spatial or temporal dimension separately, while leaves the linking of two dimensions to \pkg{cubble}. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

The rest of the paper is divided as follows: Section 2 presents the workflow of data manipulation in cubble. Section 3 explains how cubble deals with some advanced considerations  including data with hierarchical structure, data matching, and how cubble fits with existing static and interactive visualisation tools. Section 4 gives examples of the features introduced in the previous two sections with climate and hydrology data. An example on how cubble handle NetCDF data is also provided. Section 5 concludes the paper.

>>>>

Spatio-temporal data concerns variables observed in the space across time and this paper focuses on spatio-temporal vector data with fixed locations and regular time interval. This type of data can naturally been considered as a data cube with three axes: spatial identifier, temporal identifier, and variable. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows four examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

```{r illu-input, echo = FALSE, out.width = "100%", fig.cap="An illustration of different spatio-temporal data layout: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.", fig.align='center', out.height = "40%"}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

@@

None of these three formats in Figure \ref{fig:illu-input} is not particularly convenient to work with. In the single table format, to avoid plotting duplicated points of the same sites on the map, analysts always need to create a spatial variables table from the original single table. This can be a tedious step if needs to performed many times during the analysis. In another type of operation that filter the sites based on certain statistics, intermediate objects, like the per-station summary of the statistic and the filtered station ID vector, will need be created as standalone objects in the environment. The spirit of using pipe reduces the intermediate data objects that are unnecessarily created in the analysis while in this case, these only-used-once data objects can hardly be incorporated into a piping workflow in the analysis. In addition, if the data is organised as a separate spatial and temporal table as format 2 in Figure \ref{fig:illu-input}, the filter on site ID will need to performed twice in each table to keep them both up-to-date. With the array data structure, although it can easily store and wrangle large spatio-temporal data, the 3D or higher-D array can be less intuitive than a 2D tabular data structure for some analysts in data wrangling.

In the \proglang{R} community, different data objects have been proposed to work with spatio-temporal data. \pkg{spacetime} [@spacetime] is built on class \pkg{sp} [@sp] and \pkg{xts} [@xts] and both of which have been replaced by more recent spatial and temporal data objects. \pkg{spatstat} [@spatstat] implements a `ppp` class for analysing point pattern data; and more recent, \pkg{stars} [@stars] has implemented a spatio-temporal array for wrangling both vector and raster data. While these data objects are useful for its designed tasks, there has not yet been a data object that makes it easy to explore spatio-temporal data. Data objects like \pkg{sf} [@sf] and \pkg{tsibble} [@tsibble] are good examples of this in the spatial and temporal data analysis and hence there is a gap in software development for spatio-temporal data. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: climate observations from weather stations, air quality variables from air monitoring stations, or any data collected at fixed location across time.

@@

This paper describes a new tabular data object:  \pkg{cubble} for wrangling spatio-temporal data. \pkg{Cubble} overcomes the inconvenience in data wrangling mentioned above through wiring the separate spatial and temporal table into a single data object. With the new data object, analysts can spend less time organising spatial and temporal variables at different observational units and focus more on exploring the data itself. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

The rest of the paper is divided as follows: Section 2 presents the main design and functionality of \pkg{cubble}. Section 3 explains how cubble deals with more advanced considerations including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section 4 uses Australia weather station data and river level data as examples to demonstrate the use with \pkg{cubble}. An example on how \pkg{cubble} handles NetCDF data is also provided. Section 5 concludes the paper.

### 2022-02-21 main.tex
intro done

Spatio-temporal data concerns variables observed in the space across
time and this paper focuses on the spatio-temporal vector data with
fixed locations and regular time interval. This type of data can
naturally been considered as a data cube and here a cube with three
axes: spatial identifier, temporal identifier, and variable can be used
to conceptualise spatio-temporal data. In this cubic formation, space
only occupy one dimension, rather than two, to avoid the needs of
hypercube for multivariate spatio-temporal data.

While conceptually the data format is relatively simple, the actual data
object that comes to the analysts in the analysis can be in various
forms. Figure \ref{fig:illu-input} shows four examples of wild
spatio-temporal data: 1) combines the time invariant and varying
variables into a single table through duplicating the time invariant
variables; 2) splits those variables into a separate spatial table and a
temporal table; 3) \ldots; and lastly 4) is array-based data, like
NetCDF (Network Common Data Form) data, which is commonly used in earth
science, e.g.~climatology, meteorology, and oceanography, among others,
to organise multivariate data in a spatio-temporal grid.

@@

\caption[An illustration of different spatio-temporal data format]{An illustration of different spatio-temporal data format: (1) a single table for all the variables; (2) separate tables for spatial and temporal variables; (3) a table of metadata used to query the database and a separate table for queried data; and (4) an array data structure.}\label{fig:illu-input}
\end{figure}
\end{CodeChunk}

Different authors have proposed various data objects to work with
spatial temporal data in \proglang{R} and this includes: \pkg{spacetime}
\citep{spacetime} is built on the \pkg{sp} \citep{sp} and \pkg{xts}
\citep{xts} class and both of which have been replaced by more recent
spatial and temporal data objects. \pkg{spatstat} \citep{spatstat}
implements a \texttt{ppp} class for analysing point pattern data; and
more recent, \pkg{stars} \citep{stars} has implemented an
spatio-temporal array for wrangling both vector and raster data.

\textcolor{red}{Think again on the motivation and why it is hard to work existing tools.}
In spatio-temporal data analysis, analysts will receive one of the
various data formats depends on the data provider and need to manually
rearrange the data to fit into one of the data object in \proglang{R}
before starting any analysis on the data itself.

Looking at spatial or temporal data analysis, data objects like \pkg{sf}
\citep{sf} and \pkg{tsibble} \citep{tsibble} have smoothed the data
analysis in these two domains while there has not yet been a
spatio-temporal data structure that makes it easy to explore
spatio-temporal data and this creates a gap in the software development.
The requirement for such a tool is important given the ubiquity of
spatio-temporal vector data in the wild: climate observations from
weather stations, air quality variables from air monitoring stations, or
any variables that are measured at fixed location across time.

This paper describes the implementation of a new spatio-temporal data
structure: \pkg{cubble}. \pkg{Cubble} implements a data structure that
uses two forms to manage the switch between spatial and temporal
dimension. With this structure, users can manipulate the spatial or
temporal dimension separately, while leaves the linking of two
dimensions to \pkg{cubble}. The software is available from the
Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section 2 presents the
workflow of data manipulation in cubble. Section 3 explains how cubble
deals with some advanced considerations including data with hierarchical
structure, data matching, and how cubble fits with existing static and
interactive visualisation tools. Section 4 gives examples of the
features introduced in the previous two sections with climate and
hydrology data. An example on how cubble handle NetCDF data is also
provided. Section 5 concludes the paper.

@@

In cubble, data can represented in two formats: nested form and long
form, and Figure \ref{fig:illu-cubble} sketches the two forms with the
associated attributes. The decision on which form to use in data
manipulation is output-oriented, meaning analysts need to first think
about whether the output of an operation is identified only by the
spatial identifier, or a combination of spatial and temporal identifier.
The nested cubble is suitable for working with operations that are only
identified by the site and this type of operation can be a pure
manipulation of time invariant variables, or an operation that
summarises time varying variables into site. Underneath the nested form,
a cubble is built from a \code{rowwise_df} class where each site forms a
separate group. This structure simplifies the calculation that involves
temporal variables by avoiding the use of \code{purrr::map()} syntax
when working with list-column.

>>>>

Spatio-temporal data concerns variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time interval. This type of data can naturally
been considered as a data cube with three axes: spatial identifier,
temporal identifier, and variable. In this cubic formation, space only
occupies one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows four examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

@@

\caption[An illustration of different spatio-temporal data layout]{An illustration of different spatio-temporal data layout: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.}\label{fig:illu-input}
\end{figure}
\end{CodeChunk}

None of these three formats in Figure \ref{fig:illu-input} is not
particularly convenient to work with. In the single table format, to
avoid plotting duplicated points of the same sites on the map, analysts
always need to create a spatial variables table from the original single
table. This can be a tedious step if needs to performed many times
during the analysis. In another type of operation that filter the sites
based on certain statistics, intermediate objects, like the per-station
summary of the statistic and the filtered station ID vector, will need
be created as standalone objects in the environment. The spirit of using
pipe reduces the intermediate data objects that are unnecessarily
created in the analysis while in this case, these only-used-once data
objects can hardly be incorporated into a piping workflow in the
analysis. In addition, if the data is organised as a separate spatial
and temporal table as format 2 in Figure \ref{fig:illu-input}, the
filter on site ID will need to performed twice in each table to keep
them both up-to-date. With the array data structure, although it can
easily store and wrangle large spatio-temporal data, the 3D or higher-D
array can be less intuitive than a 2D tabular data structure for some
analysts in data wrangling.

In the \proglang{R} community, different data objects have been proposed
to work with spatio-temporal data. \pkg{spacetime} \citep{spacetime} is
built on class \pkg{sp} \citep{sp} and \pkg{xts} \citep{xts} and both of
which have been replaced by more recent spatial and temporal data
objects. \pkg{spatstat} \citep{spatstat} implements a \texttt{ppp} class
for analysing point pattern data; and more recent, \pkg{stars}
\citep{stars} has implemented a spatio-temporal array for wrangling both
vector and raster data. While these data objects are useful for its
designed tasks, there has not yet been a data object that makes it easy
to explore spatio-temporal data. Data objects like \pkg{sf} \citep{sf}
and \pkg{tsibble} \citep{tsibble} are good examples of this in the
spatial and temporal data analysis and hence there is a gap in software
development for spatio-temporal data. The requirement for such a tool is
important given the ubiquity of spatio-temporal vector data in the wild:
climate observations from weather stations, air quality variables from
air monitoring stations, or any data collected at fixed location across
time.

This paper describes a new tabular data object: \pkg{cubble} for
wrangling spatio-temporal data. \pkg{Cubble} overcomes the inconvenience
in data wrangling mentioned above through wiring the separate spatial
and temporal table into a single data object. With the new data object,
analysts can spend less time organising spatial and temporal variables
at different observational units and focus more on exploring the data
itself. The software is available from the Comprehensive R Archive
Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section 2 presents the main
design and functionality of \pkg{cubble}. Section 3 explains how cubble
deals with more advanced considerations including data with hierarchical
structure, data matching, how cubble fits with existing static and
interactive visualisation tools, and spatio-temporal data
transformation. Section 4 uses Australia weather station data and river
level data as examples to demonstrate the use with \pkg{cubble}. An
example on how \pkg{cubble} handles NetCDF data is also provided.
Section 5 concludes the paper.

@@

In cubble, a single data object can be arranged into two forms: nested
form and long form, and Figure \ref{fig:illu-cubble} sketches the two
forms with the associated attributes. The decision on which form to use
in data wrangling is output-oriented, meaning analysts need to first
think about whether the output of an operation is identified only by the
spatial identifier, or a combination of spatial and temporal identifier.
The nested cubble is suitable for working with operations that are only
identified by the site and this type of operation can be a pure
manipulation of time invariant variables, or an operation that
summarises time varying variables into site. Underneath the nested form,
a cubble is built from a \code{rowwise_df} class where each site forms a
separate group. This structure simplifies the calculation that involves
temporal variables by avoiding the use of \code{purrr::map()} syntax
when working with list-column.
